A Stopped Negative Binomial Distribution

### 2016-08-22 stat-and-prob-letters/snb-paper.tex
Paper submitted.

\modulolinenumbers[5]

@@

\linenumbers

@@

If all 17 patients are enrolled at once, as in the classic
design, then the sample size is 17. However, in most clinical trials the
patients are enrolled sequentially over time.
In the present example, observing seven
successful patients is one endpoint and the number of enrollees required
could be as small as seven. Similarly 11
observed treatment failures also ends the trial. This sampling mechanism, in
which the experiment ends as soon as any predefined endpoint is reached, is
called {\em curtailed sampling}. Under curtailed sampling, the range of the
sample size for this trial is seven through 17.

@@

\begin{figure}[bp!]
\includegraphics[width=\textwidth]{KanePlot.pdf}
\caption{
A hypothetical realization of the trial.
}
\label{fig:kane_viz}
\end{figure}

@@

\begin{figure}[bp!]
\centering
\subfloat[enrollment distribution]{\includegraphics[width=0.5\textwidth]{snb-first-plot.pdf}}
\hfill
\subfloat[mean and variance varying $p$]{\includegraphics[width=0.5\textwidth]{mean-and-variance.pdf}}
\caption{
The distribution of a trial that stops after seven patients respond to treatment
or 11 patients do not. Panel (a) shows the distribution when
$p = 0.2$. The mean and variance of the distribution varying $p$ 
between zero and one are shown in panel (b).
}
\label{fig:exp-and-var}
\end{figure}

@@

To show that (\ref{eqn:pmf}) sums to one, define
%\begin{align} \label{eqn:sum_proof}
%R &= \sum_{k=s}^{s+t-1} S(k, p, s) + \sum_{k=t}^{s+t-1} S(k, 1-p, t) \\
%  &= \sum_{k=s}^{s+t-1} {k-1 \choose s-1} p^s (1-p)^{k-s} + \sum_{k=t}^{s+t-1} {k-1 \choose k-t} p^{k-t} (1-p)^t
%\end{align}
\begin{equation*} 
R = \sum_{k=s}^{s+t-1} S(k, p, s) + \sum_{k=t}^{s+t-1} S(k, 1-p, t).
\end{equation*}
If we substitute $i=k-s$ in the first summation and $j=k-t$ in the second then
$R$ can be written as the cumulative distribution function of two
negative binomial distributions:
\begin{equation} \label{eqn:transformed_sum}
R = \sum_{i=0}^{t-1} {i+s-1 \choose i} p^s (1-p)^i \; + \;
\sum_{j=0}^{s-1} {j+t-1 \choose j} p^j (1-p)^t.
\end{equation}

Let $\mathcal{I}_p(s, t)$ be the {\em regularized incomplete beta function} 
\citep{Olver2010}. This function satisfies 
$\mathcal{I}_p(s, t) = 1-\mathcal{I}_{1-p}(t, s)$ \citep{Uppuluri1967}. Then
\begin{align*}
R = \sum_{i=0}^{t-1} &{i+s-1 \choose i} p^s (1-p)^i +
\sum_{j=0}^{s-1}  {j+t-1 \choose j} p^j  (1-p)^t \\
   &= 1-\mathcal{I}_p(s, t) + 1 - \mathcal{I}_{1-p}(t, s) \\
   &= 1. 
\end{align*}
This completes the proof that (\ref{eqn:pmf}) is the distribution of the
stopping time and is a valid probability mass function.
\end{proof}

@@

\begin{figure}[p!]
\begin{center}
\includegraphics[width=\textwidth]{shapes.pdf}
\end{center}
\caption{Different shapes of the SNB distribution with parameters 
($p$, $s$, $t$), as given. Black indicates mass contributed by reaching
$s$ responders before $t$ non-responders. Grey indicates
mass contributed by reaching $t$ non-responders first. \label{shapes.fig}}
\end{figure}

@@

The SNB generalizes both the minimum (riff-shuffle) and maximum negative
binomial distributions up to a translation of the support.
For the special case of $\,s=t,$ the distribution of $\,Y\,$ is the
riff-shuffle, or minimum negative binomial 
distribution~\citep{Uppuluri1967,Johnson2005}.
The maximum negative binomial \cite{Zhang2000,Zelterman2005,Johnson2005} is 
the smallest number of outcomes necessary to 
observe at least $s$ responders {\em and} $s$ non-responders and is equivalent
to a translated version of the riff-shuffle.

@@

Drawing connections to the gamma and Poisson distributions are more complicated.
However a connection to the gamma distribution well-studied problem in the 
literature (see \cite{Ord1968,Guenther1972,Best1974} for examples). 
A connection to the Poisson appears in \cite{Anscombe1950} where
it is shown that if the mean of a Poisson is proportional
to a chi-square distribution with $2k$ degrees of freedom
then the negative binomial is obtained. 
Both of these approximations work by equating cumulants and then showing 
that differences between the cumulant generating
functions converge to zero.

>>>>

%\modulolinenumbers[5]

@@

%\linenumbers

@@

If all 17 patients are enrolled at once, as in the classic
design, then the sample size is 17. However, in most clinical trials the
patients are enrolled sequentially over time.
In the present example, observing seven
successful patients ends the trial and so the number of enrollees required
could be as small as seven. Similarly 11
observed treatment failures also ends the trial. This sampling mechanism, in
which the experiment ends as soon as any predefined endpoint is reached, is
called {\em curtailed sampling}. Under curtailed sampling, the range of the
sample size for this trial is seven through 17.

@@

\begin{figure}[bp!]
\centering
\includegraphics[width=0.9\textwidth]{KanePlot.pdf}
\caption{
A hypothetical realization of the trial.
}
\label{fig:kane_viz}
\end{figure}

@@

\begin{figure}[bp!]
\centering
\subfloat[enrollment distribution]{\includegraphics[width=0.45\textwidth]{snb-first-plot.pdf}}
\hfill
\subfloat[mean and variance varying $p$]{\includegraphics[width=0.45\textwidth]{mean-and-variance.pdf}}
\caption{
The distribution of the sample size in a trial that stops after seven 
patients respond to treatment
or 11 patients do not. Panel (a) shows the distribution when
$p = 0.2$. The mean and variance of the distribution varying $p$ 
between zero and one are shown in panel (b).
}
\label{fig:exp-and-var}
\end{figure}

@@

To show that (\ref{eqn:pmf}) sums to one, define
%\begin{align} \label{eqn:sum_proof}
%R &= \sum_{k=s}^{s+t-1} S(k, p, s) + \sum_{k=t}^{s+t-1} S(k, 1-p, t) \\
%  &= \sum_{k=s}^{s+t-1} {k-1 \choose s-1} p^s (1-p)^{k-s} + \sum_{k=t}^{s+t-1} {k-1 \choose k-t} p^{k-t} (1-p)^t
%\end{align}
\begin{equation*} 
R = \sum_{k=s}^{s+t-1} S(k, p, s) + \sum_{k=t}^{s+t-1} S(k, 1-p, t).
\end{equation*}
If we substitute $i=k-s$ in the first summation and $j=k-t$ in the second then
$R$ can be written as the cumulative distribution function (CDF) of two
negative binomial distributions:
\begin{equation} \label{eqn:transformed_sum}
R = \sum_{i=0}^{t-1} {i+s-1 \choose i} p^s (1-p)^i \; + \;
\sum_{j=0}^{s-1} {j+t-1 \choose j} p^j (1-p)^t.
\end{equation}

The CDF of the negative binomial is commonly expressed in terms of the
{\em regularized incomplete beta function} \citep{Olver2010}. Using standard
notation, the first summation in (\ref{eqn:transformed_sum})
is written as $\mathcal{I}_{1-p}(t, s)$.
This function satisfies 
$\mathcal{I}_p(s, t) = 1-\mathcal{I}_{1-p}(t, s)$ \citep{Uppuluri1967}. Then
\begin{align*}
R = \sum_{i=0}^{t-1} &{i+s-1 \choose i} p^s (1-p)^i +
\sum_{j=0}^{s-1}  {j+t-1 \choose j} p^j  (1-p)^t \\
   &= 1-\mathcal{I}_p(s, t) + 1 - \mathcal{I}_{1-p}(t, s) \\
   &= 1. 
\end{align*}
This completes the proof that (\ref{eqn:pmf}) is the distribution of the
stopping time and is a valid probability mass function.
\end{proof}

@@

\begin{figure}[p!]
\begin{center}
\includegraphics[width=0.8\textwidth]{shapes.pdf}
\end{center}
\caption{Different shapes of the SNB distribution with parameters 
($p$, $s$, $t$), as given. Black indicates mass contributed by reaching
$s$ responders before $t$ non-responders. Grey indicates
mass contributed by reaching $t$ non-responders first. \label{shapes.fig}}
\end{figure}

@@

The SNB generalizes both the minimum (riff-shuffle) and maximum negative
binomial distributions up to a translation of the support.
For the special case of $\,s=t,$ the distribution of $\,Y\,$ is the
riff-shuffle, or minimum negative binomial 
distribution~\citep{Uppuluri1967,Johnson2005}.
The maximum negative binomial \cite{Johnson2005,Zhang2000,Zelterman2005} is 
the smallest number of outcomes necessary to 
observe at least $s$ responders {\em and} $s$ non-responders and is equivalent
to a translated version of the riff-shuffle.

@@

Drawing connections to the gamma and Poisson distributions are more complicated.
However a connection to the gamma distribution well-studied problem in the 
literature (see \cite{Best1974,Ord1968,Guenther1972} for examples). 
A connection to the Poisson appears in \cite{Anscombe1950} where
it is shown that if the mean of a Poisson is proportional
to a chi-square distribution with $2k$ degrees of freedom
then the negative binomial is obtained. 
Both of these approximations work by equating cumulants and then showing 
that differences between the cumulant generating
functions converge to zero.
